---
title: "Example WNV simulation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example WNV simulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(MicroWNV)
library(ggplot2)
library(data.table)
library(parallel)
```

In MicroWNV, the goal is to get various models fill in a set of component interfaces; if you want to write a new aquatic model, you fill in the interface and then plug into the existing framework, and let the simulation handle the rest! Here we present the most simple case of a Ross-Macdonald mosquito model linked to a trace derived (exogeneously forced) model of emergence, 
a SIRS (Susceptible-Infectious-Recovered-Susceptible) adult bird model lined to a forced model of fledglings (immature birds),
and a SIR (Susceptible-Infectious-Recovered) human model.

In this model there is only 1 patch and 1 human strata. We set up the model parameters
here. We let mosquito emergence be seasonal, so $\lambda$ varies with time, peaking
in the "middle" of the year with 50 mosquitoes emerging per day. 
We will run the simulation for 3 years.

We set up the birds with an offset signal for birth of fledglings peaking at 5 fledglings
emerging per day at the start of the year.

Please note that these specific parameter values have not been calibrated nor are
representative of WNV, but instead chosen to illustrate interesting dynamics.

```{R}
patches <- 1
n <- 1
tmax <- 3 * 365

# human parameters
theta <- diag(n)
SIR <- matrix(data = c(1e3, 0, 0), nrow = 1, ncol = 3)
gamma <- 1/7

# bird parameters
theta_B <- diag(patches)
SIR_B <- matrix(data = c(9e3, 1e3, 0), nrow = 1, ncol = 3)
gammaB <- 1/20
rB <- 1/100
mu <- 1/365

# mosquito parameters
f <- 0.3
q <- 0.0001
eip <- 10
p <- 0.95
M <- 500
Y <- 50
Z <- 5
psi <- diag(patches)

lambda <- dnorm(x = 1:365, mean = 180, sd = 90)
lambda <- lambda * (5e3/max(lambda))
lambda <- t(replicate(patches, lambda))

fledge_lambda <- dnorm(x = c(182:365, 365:185), mean = 180, sd = 90)
fledge_lambda <- fledge_lambda * (50/max(fledge_lambda))
fledge_lambda <- t(replicate(patches, fledge_lambda))
```

## MicroWNV simulation

Now we set up the model and run a deterministic simulation.

```{R}
mod <- make_microWNV(tmax = tmax, p = patches)

# humans
setup_humans_SIR(mod, stochastic = FALSE, theta = theta, H = sum(SIR), SIR = SIR, gamma = gamma)

# birds
setup_birds_SIRS(mod, stochastic = FALSE, fledge_disperse = theta_B, theta = theta_B, SIR = SIR_B, mu = mu, gamma = gammaB, r = rB)
setup_fledge_trace(mod, stochastic = FALSE, trace = fledge_lambda)

# mosquitoes
setup_mosquito_RM(mod, stochastic = FALSE, f = f, q = q, eip = eip, p = p, psi = psi, M = M, Y = Y, Z = Z)
setup_aqua_trace(mod, stochastic = FALSE, lambda = lambda)

# output
det_out <- rbind(
  data.table(Day = rep(1:tmax, times = 3), Count = NaN, Compartment = rep(c("S", "I", "R"), each = tmax), Species = "Human"),
  data.table(Day = rep(1:tmax, times = 3), Count = NaN, Compartment = rep(c("S", "I", "R"), each = tmax), Species = "Bird"),
  data.table(Day = rep(1:tmax, times = 3), Count = NaN, Compartment = rep(c("M", "Y", "Z"), each = tmax), Species = "Mosquito")
)
setkeyv(det_out, c("Day", "Species"))

# run it
while (mod$global$tnow <= tmax) {
  # compute biting
  compute_bloodmeal(model = mod)
  # update state
  step_aqua(model = mod)
  step_mosquitoes(model = mod)
  step_fledge(model = mod)
  step_birds(model = mod)
  step_humans(model = mod)
  # output
  det_out[Day == mod$global$tnow  & Compartment %in% c("S", "I", "R") & Species == "Human", "Count" := as.vector(mod$human$SIR)]
  det_out[Day == mod$global$tnow & Compartment %in% c("S", "I", "R") & Species == "Bird", "Count" := as.vector(mod$bird$SIR)]
  det_out[Day == mod$global$tnow & Compartment %in% c("M", "Y", "Z") & Species == "Mosquito", "Count" := c(mod$mosquito$M, mod$mosquito$Y, mod$mosquito$Z)]
  # tick
  mod$global$tnow <- mod$global$tnow + 1L
}
```

Now we draw 10 trajectories from the stochastic simulation, and plot output. We plot the cloud of stochastic trajectories as faint lines and the deterministic solution as solid lines.

```{R}
sto_out <- mclapply(X = 1:10, FUN = function(runid) {
  mod <- make_microWNV(tmax = tmax, p = patches)

  setup_humans_SIR(mod, stochastic = TRUE, theta = theta, H = sum(SIR), SIR = SIR, gamma = gamma)
  setup_birds_SIRS(mod, stochastic = TRUE, fledge_disperse = theta_B, theta = theta_B, SIR = SIR_B, mu = mu, gamma = gammaB, r = rB)
  setup_fledge_trace(mod, stochastic = TRUE, trace = fledge_lambda)
  setup_mosquito_RM(mod, stochastic = TRUE, f = f, q = q, eip = eip, p = p, psi = psi, M = M, Y = Y, Z = Z)
  setup_aqua_trace(mod, stochastic = TRUE, lambda = lambda)
  
  # output
  out <- rbind(
    data.table(Day = rep(1:tmax, times = 3), Count = NaN, Compartment = rep(c("S", "I", "R"), each = tmax), Species = "Human"),
    data.table(Day = rep(1:tmax, times = 3), Count = NaN, Compartment = rep(c("S", "I", "R"), each = tmax), Species = "Bird"),
    data.table(Day = rep(1:tmax, times = 3), Count = NaN, Compartment = rep(c("M", "Y", "Z"), each = tmax), Species = "Mosquito")
  )
  setkeyv(out, c("Day", "Species"))
  
  # run it
  while (mod$global$tnow <= tmax) {
    # compute biting
    compute_bloodmeal(model = mod)
    # update state
    step_aqua(model = mod)
    step_mosquitoes(model = mod)
    step_fledge(model = mod)
    step_birds(model = mod)
    step_humans(model = mod)
    # output
    out[Day == mod$global$tnow  & Compartment %in% c("S", "I", "R") & Species == "Human", "Count" := as.vector(mod$human$SIR)]
    out[Day == mod$global$tnow & Compartment %in% c("S", "I", "R") & Species == "Bird", "Count" := as.vector(mod$bird$SIR)]
    out[Day == mod$global$tnow & Compartment %in% c("M", "Y", "Z") & Species == "Mosquito", "Count" := c(mod$mosquito$M, mod$mosquito$Y, mod$mosquito$Z)]
    # tick
    mod$global$tnow <- mod$global$tnow + 1L
  }
  
  out[, "Run" := as.integer(runid)]
  return(out)
})

sto_out <- do.call(rbind, sto_out)
```

The results are plotted below.

```{R}
ggplot(sto_out) +
    geom_line(aes(x = Day, y = Count, color = Compartment, group = interaction(Run, Compartment)), alpha = 0.3) +
    geom_line(data = det_out, aes(x = Day, y = Count, color = Compartment)) +
    facet_wrap(. ~ Species, scales = "free")
```
